---
title: "Amyloid"
author: "Liwen Yin"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

## data cleaning

```{r}
library(dplyr)
neg <- read.csv("Amyloid_negative.csv")
pos <- read.csv("Amyloid_positive.csv")
amyloid <- rbind(neg, pos) 
colnames(amyloid)[which(names(amyloid) == "Amyloid")] <- "amyloid_status"

amyloid <- amyloid %>%
  dplyr::select(-matches("X")) %>%
  mutate(amyloid_status = factor(ifelse(amyloid_status == "Y", 1, 0))) %>%
  mutate(
    Laterality = as.factor(Laterality),
    Race = as.factor(Race),
    Monoclonal.Gammopathy = as.factor(Monoclonal.Gammopathy),
    Rheumatoid.Arthritis = as.factor(Rheumatoid.Arthritis),
    Coronary.Artery.Disease = as.factor(Coronary.Artery.Disease),
    Afib = as.factor(Afib),
    Degenerative.Spine.Disease = as.factor(Degenerative.Spine.Disease),
    Diabetes = as.factor(Diabetes),
    Tendinopathy = as.factor(Tendinopathy),
    EMG = as.factor(EMG),
    Bilateral = as.factor(Bilateral.)
  )
amyloid$Grade <- factor(amyloid$Grade, 
                   levels = c(0, "mild", "moderate", "severe"), 
                   labels = c("none", "mild", "moderate", "severe"))
amyloid<- amyloid %>% filter(!is.na(Grade))
amyloid <- amyloid %>%
  mutate(
    Severity = factor(
      ifelse(Grade %in% c("none", "mild", "moderate"), "Low", "Severe"),
      levels = c("Low", "Severe") 
    )
  )
write.csv(amyloid, "amyloid.csv", row.names = FALSE)
```

## logistic regression model

```{r}
library(rstanarm)
library(arm)
library(pROC)
model1 <- glm(amyloid_status ~ Race + Age + Afib + Degenerative.Spine.Disease + Bilateral. + Monoclonal.Gammopathy + Rheumatoid.Arthritis + Severity, data = amyloid, family = binomial)
summary(model1)
set.seed(1)
predicted_probs <- predict(model1, type = "response")
roc_curve <- roc(amyloid$amyloid_status, predicted_probs)
auc_value <- auc(roc_curve)
plot(roc_curve,  main = paste("ROC Curve for glm Model (AUC =", round(auc_value, 3), ")"), 
     col = "blue", 
     lwd = 2, legacy.axes = TRUE)
print(paste("AUC:", round(auc_value, 3)))
```

## Bayesian generalized linear models
```{r}
library(rstanarm)
library(arm)
library(pROC)
bayes_model <- stan_glm(amyloid_status ~ Grade, 
                        data = amyloid, 
                        family = binomial, 
                        prior = student_t(3, 0, 2.5),  #weakly informative
                        chains = 4, iter = 2000, seed = 123,refresh = 0)

summary(bayes_model)
#Credible Interval
posterior_interval(bayes_model, prob = 0.95)
#pp check
pp_check(bayes_model)
#Odds ratio
exp(posterior_interval(bayes_model, prob = 0.95))
#ROC & AUC
set.seed(1)
predicted_probs <- posterior_predict(bayes_model, type = "response") %>% colMeans()
actual_values <- amyloid$amyloid_status
roc_curve <- roc(actual_values, predicted_probs,levels = c(0, 1))
auc_value <- auc(roc_curve)
plot(roc_curve,  main = paste("ROC Curve for Bayes Model (AUC =", round(auc_value, 3), ")"), 
     col = "blue", 
     lwd = 2, legacy.axes = TRUE)
print(paste("AUC:", round(auc_value, 3)))
```

For every unit increase in the linear component of Grade, the log-odds of amyloid positivity increase by 3.2.
mean_posterior predictive distribution:the mean is 0.2, suggesting the model predicts an average probability of positive as approximately 20% across the dataset.
95% Bayesian credible intervals : The linear effect of Grade is strongly positive.
The quadratic effect of Grade includes zero, suggesting no significant quadratic effect.
odds ratio: Linear Effect of Grade OR: [0.01, 0.21] : The baseline odds of amyloid positivity are very low.
Quadratic Effect of Grade OR: [2.51, 1211.14] : Indicates that higher Grade strongly increases the odds of amyloid positivity.
pp check: Posterior predictive distribution compared to the observed data distribution; Slight deviations at very low probabilities.

```{r}
bayes_model2 <- stan_glm(
  amyloid_status ~ Grade, 
  data = amyloid, 
  family = binomial,
  prior = normal(0, 1, autoscale = TRUE),  # Ordered effect prior
  chains = 4, iter = 2000, seed = 123,refresh = 0
)
summary(bayes_model2)
set.seed(1)
predicted_probs <- posterior_predict(bayes_model2, type = "response") %>% colMeans()
actual_values <- amyloid$amyloid_status
roc_curve <- roc(actual_values, predicted_probs,levels = c(0, 1))
auc_value <- auc(roc_curve)
plot(roc_curve,  main = paste("ROC Curve for Severity Bayes Model (AUC =", round(auc_value, 3), ")"), 
     col = "blue", 
     lwd = 2, legacy.axes = TRUE)
print(paste("AUC:", round(auc_value, 3)))
```

Autoscaling adjusts the prior standard deviation relative to the data scale, ensuring that the prior matches the actual predictor variability.
The normal prior with scaling ensures coefficients are regularized to prevent extreme estimates in case of sparse data.

```{r}
library(rstanarm)
bayes_model3 <- stan_glm(
  amyloid_status ~ Race + Age + Afib + Degenerative.Spine.Disease + Bilateral. + Monoclonal.Gammopathy + Rheumatoid.Arthritis +Severity,
  data = amyloid,
  family = binomial,
 prior = normal(0, 5),  
  chains = 4, iter = 2000, seed = 123, refresh = 0
)
print(summary(bayes_model3), digits=4)
set.seed(1)
predicted_probs <- posterior_predict(bayes_model3, type = "response") %>% colMeans()
actual_values <- amyloid$amyloid_status
roc_curve <- roc(actual_values, predicted_probs,levels = c(0, 1))
auc_value <- auc(roc_curve)
plot(roc_curve,  main = paste("ROC Curve for Bayes Model 3 (AUC =", round(auc_value, 3), ")"), 
     col = "blue", 
     lwd = 2, legacy.axes = TRUE)
print(paste("AUC:", round(auc_value, 3)))
```

```{r}
library(rstanarm)
bayes_model6 <- stan_glm(
  amyloid_status ~ Race + Age + Afib + Degenerative.Spine.Disease + Bilateral. + Monoclonal.Gammopathy + Rheumatoid.Arthritis +Severity,
  data = amyloid,
  family = binomial,
 prior = student_t(1,0, 5),  
  chains = 4, iter = 2000, seed = 123, refresh = 0
)
print(summary(bayes_model6), digits=4)
set.seed(1)
predicted_probs <- posterior_predict(bayes_model6, type = "response") %>% colMeans()
actual_values <- amyloid$amyloid_status
roc_curve <- roc(actual_values, predicted_probs,levels = c(0, 1))
auc_value <- auc(roc_curve)
plot(roc_curve,  main = paste("ROC Curve for Bayes Model 6 (AUC =", round(auc_value, 3), ")"), 
     col = "blue", 
     lwd = 2, legacy.axes = TRUE)
print(paste("AUC:", round(auc_value, 3)))
```


normal(0, 1) indicates no strong prior belief about the direction (positive or negative) of the effect of the predictors on the log-odds of amyloid_status.
A standard deviation of 1 allows for moderate variability in the coefficients.
prior_intercept = normal(0, 5):Allows for wide baseline probabilities, accommodating substantial uncertainty in the initial prevalence of the outcome.

```{r}
bayes_model4 <- stan_glm(
  amyloid_status ~ Points,  
  data = amyloid,          
  family = binomial,        
  prior = normal(0, 1),     # Weakly informative
  prior_intercept = normal(0, 5),  
  chains = 4, iter = 2000, seed = 123, refresh = 0
)
summary(bayes_model4)
set.seed(1)
predicted_probs <- posterior_predict(bayes_model4, type = "response") %>% colMeans()
actual_values <- amyloid$amyloid_status
roc_curve <- roc(actual_values, predicted_probs,levels = c(0, 1))
auc_value <- auc(roc_curve)
plot(roc_curve,  main = paste("ROC Curve for Bayes Model 2 (AUC =", round(auc_value, 3), ")"), 
     col = "blue", 
     lwd = 2, legacy.axes = TRUE)
print(paste("AUC:", round(auc_value, 3)))
pp_check(bayes_model4)
```

```{r}
reg <- glm(amyloid_status ~ Points, family = binomial, data = amyloid)
summary(reg)
library(pROC)
set.seed(1)
predicted_probs <- predict(reg, type = "response")
actual_values <- amyloid$amyloid_status
roc_curve <- roc(actual_values, predicted_probs)
auc_value <- auc(roc_curve)
auc_value_rounded <- round(auc_value, 3)
plot(roc_curve, main = paste("ROC Curve for Logistic Regression Model (AUC:", auc_value_rounded, ")"), col = "blue", lwd = 2, legacy.axes = TRUE)
print(paste("AUC=", auc_value_rounded))
```

If the effect of Points is relatively small (statistically significant in glm, but with a small actual effect size), the prior distribution tends to shrink the coefficient of Points toward 0.

amyloid_status ~ Points + grade
```{r}
bayes_model5 <- stan_glm(
  amyloid_status ~ Grade + Points,  
  data = amyloid,          
  family = binomial,        
  prior = normal(0, 5),     # Weakly informative
  prior_intercept = normal(0, 5),  
  chains = 4, iter = 2000, seed = 123, refresh = 0
)
print(summary(bayes_model5),digit = 4)
library(pROC)
set.seed(1)
predicted_probs <- predict(bayes_model5, type = "response")
actual_values <- amyloid$amyloid_status
roc_curve <- roc(actual_values, predicted_probs)
auc_value <- auc(roc_curve)
auc_value_rounded <- round(auc_value, 3)
plot(roc_curve, main = paste("ROC Curve for Bayes Model (AUC=", auc_value_rounded, ")"), col = "blue", lwd = 2, legacy.axes = TRUE)
print(paste("AUC=", auc_value_rounded))
```

