---
title: "Exploratory Data Analysis on Enhancers Project"
author: "Liwen Yin"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---
```{r}
knitr::opts_chunk$set(echo = FALSE)
```
## EDA 
# Distribution of activity score in three treatment groups
Each plot shows a density histogram with an overlaid smoothed density curve. All groups show a right-skewed distribution, which means most enhancers have low activity scores, with a smaller number showing very high activity.
```{r}
library(dplyr)

setwd("C:/Users/annik/OneDrive/Desktop/BU Fall 2024/MA675/Genetics Project")
hksm <- read.csv("EG_HKSM_20E_col_03_11.csv")
e20 <- read.csv("EG_20E_col_03_11.csv")
con <- read.csv("EG_Con_col_03_11.csv")
hksm <- hksm %>% select(-Treatment)
```

```{r}
library(dplyr)
library(ggplot2)
plot_new_act_score <- function(data, dataset_name) {
  ggplot(data, aes(x = new_act_score)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black", alpha = 0.7) +
    geom_density(color = "red", size = 1) +
    labs(title = paste("Distribution of new_act_score in", dataset_name),
         x = "new_act_score",
         y = "Density") +
    theme_minimal()
}
#control
plot_new_act_score(con, "control group")
#e20 
plot_new_act_score(e20, "e20 group")
#hksm
plot_new_act_score(hksm, "hksm group")
```

```{r}
library(dplyr)
#special column
cols_con <- names(con)
cols_e20 <- names(e20)
cols_hksm <- names(hksm)
con_only <- setdiff(cols_con, union(cols_e20, cols_hksm))
e20_only <- setdiff(cols_e20, union(cols_con, cols_hksm))
hksm_only <- setdiff(cols_hksm, union(cols_con, cols_e20))
#cat(" con:\n"); print(con_only)
#cat("\n e20:\n"); print(e20_only)
#cat("\n hksm:\n"); print(hksm_only)
#combine by common column
con <- con %>% mutate(treatment = "H2O")
hksm <- hksm %>% mutate(treatment = "HKSM")
e20 <- e20 %>% mutate(treatment = "20E")
common_cols <- Reduce(intersect, list(names(con), names(hksm), names(e20)))
con_common <- con %>% select(all_of(common_cols))
hksm_common <- hksm %>% select(all_of(common_cols))
e20_common <- e20 %>% select(all_of(common_cols))
cdata <- bind_rows(con_common, hksm_common, e20_common) %>%
  mutate(treatment = factor(treatment, levels = c("H2O", "HKSM", "20E"))) %>% 
  select(-Activity_score)
```
#if they share the same enhancer--no same enhancer among 3 groups
```{r}
enhancer_con <- unique(con$Enhancer)
enhancer_e20 <- unique(e20$Enhancer)
enhancer_hksm <- unique(hksm$Enhancer)
common_enhancers <- Reduce(intersect, list(enhancer_con, enhancer_e20, enhancer_hksm))
length(common_enhancers)
head(common_enhancers)
```
#Here is the boxplot. The box represents the interquartile range (IQR), which includes the middle 50% of the data. The dots outside the whiskers represent outliers, values that fall outside 1.5 times the IQR. These are higher activity scores that deviate from the general distribution.
```{r}
#boxplot
library(ggplot2)
ggplot(cdata, aes(x = treatment, y = new_act_score, fill = treatment)) +
  geom_boxplot() +
  labs(
    title = "Enhancer Activity Score by Treatment",
    x = "Treatment Group",
    y = "Enhancer Activity Score (new_act_score)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
#TF enrichment analysis
We are wondering which transcription factors (TFs) are more common in high activity enhancers(activity score over 1000)?
The bar plot shown below represents the fold enrichment of transcription factors (TFs) in high-activity enhancers for three treatment groups. It displays the top enriched TFs, comparing their frequency in high-activity enhancers versus the entire dataset.
$$\text{Fold Enrichment} = \frac{\text{Mean TF motif count in high-activity enhancers}}{\text{Mean TF motif count in all enhancers}}$$

```{r}
library(dplyr)
library(ggplot2)

non_tf_cols <- c("Enhancer", "Genes", "new_act_score", "treatment", "TBS", "Length")
tf_cols <- setdiff(names(cdata), non_tf_cols)

plot_tf_enrichment <- function(df, treatment_label, threshold = 1000) {
  df_t <- df %>% filter(treatment == treatment_label)
  high_df <- df_t %>% filter(new_act_score > threshold)
  tf_means_all <- colMeans(df_t[, tf_cols], na.rm = TRUE)
  tf_means_high <- colMeans(high_df[, tf_cols], na.rm = TRUE)
  tf_enrichment <- data.frame(
    TF = tf_cols,
    Mean_All = tf_means_all,
    Mean_High = tf_means_high
  ) %>%
    mutate(Fold_Enrichment = Mean_High / (Mean_All + 1e-6)) %>%
    arrange(desc(Fold_Enrichment)) %>%
    head(15)

  ggplot(tf_enrichment, aes(x = reorder(TF, Fold_Enrichment), y = Fold_Enrichment)) +
    geom_col(fill = "tomato") +
    coord_flip() +
    labs(
      title = paste("Top Enriched TFs in High-Activity Enhancers (", treatment_label, ")", sep = ""),
      x = "Transcription Factor (TF)",
      y = "Fold Enrichment (High vs All)"
    ) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
    theme_minimal()
}

plot_tf_enrichment(cdata, "H2O")
plot_tf_enrichment(cdata, "HKSM")
plot_tf_enrichment(cdata, "20E")
```
The higher the TF ranking, the more likely it is to appear on enhancers with higher activity
A Fold Enrichment > 1 means the TF is more common in high-activity enhancers than in general  potentially important for enhancer activity.
A Fold Enrichment ≈ 1 means no real difference.
A Fold Enrichment < 1 means the TF is less frequent in high-activity enhancers.



